---
title: Trajectory utilities for package developers
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: 8 November 2020"
output:
  BiocStyle::html_document:
    toc_float: true
package: TrajectoryUtils
vignette: >
  %\VignetteIndexEntry{Trajectory utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

The `r Biocpkg("TrajectoryUtils")` package contains low-level utilities to support trajectory inference packages.
_Support_ is the key word here: this package does not perform any inference itself, deferring that to higher-level packages like `r Biocpkg("TSCAN")` and `r Biocpkg("slingshot")`.
In fact, most of the code in this package was factored out of those two packages after we realized their commonalities.
By rolling this out into a separate package, we can cut down on redundancy, facilitate better propagation of features and simplify maintenance.
If you're a developer and you have a general utility for trajectory inference, contributions are welcome.

# MST construction

The construction of a cluster-based minium spanning tree is the primary motivator for this package, 
as it was implemented separately in `r Biocpkg("TSCAN")` and `r Biocpkg("slingshot")`.
The idea is simple - cluster cells and create a minimum spanning tree from the cluster centroids to serve as the backbone for the trajectory reconstruction.

```{r}
# Mocking up a Y-shaped trajectory.
centers <- rbind(c(0,0), c(0, -1), c(1, 1), c(-1, 1))
rownames(centers) <- seq_len(nrow(centers))
clusters <- sample(nrow(centers), 1000, replace=TRUE)
cells <- centers[clusters,]
cells <- cells + rnorm(length(cells), sd=0.5)

# Creating the MST:
library(TrajectoryUtils)
mst <- createClusterMST(cells, clusters)
plot(mst)
```

The implementation in `createClusterMST()` combines several useful ideas from the two aforementioned packages:

- With `outgroup=TRUE`, users can add an outgroup to break apart distant clusters.
This ensures that the MST does not form spurious links between unrelated parts of the dataset.
- With `with.mnn=TRUE`, users can create the MST based on the distances between mutually nearest neighbors.
This avoids penalizing the formation of edges between adjacent heterogeneous clusters. 
- With `use.median=TRUE`, the centroids are computed by taking the median instead of the mean.
This protects against clusters with many outliers.

# The `PseudotimeOrdering` class

We can create a compact representation of pseudotime orderings in the form of a matrix where rows are cells and columns are paths through the trajectory.
Each cell receives a pseudotime value along a path - or `NA`, if the cell does not lie on that path.
On occasion, we may wish to annotate this with metadata on the cells or on the paths.
This is supported by the `PseudotimeOrdering` class:

```{r}
# Make up a matrix of pseudotime orderings.
ncells <- 200
npaths <- 5
orderings <- matrix(rnorm(1000), ncells, npaths)

# Default constructor:
(pto <- PseudotimeOrdering(orderings))
```

It is then straightforward to add metadata on the cells:

```{r}
pto$cluster <- sample(LETTERS[1:5], ncells, replace=TRUE)
cellData(pto)
```

Or on the paths:

```{r}
pathData(pto)$description <- sprintf("PATH-%i", seq_len(npaths))
pathData(pto)
```

Additional matrices can also be added if multiple statistics are available for each cell/path combination.
For example, one might imagine assigning the confidence to which each cell is assigned to each path.

```{r}
pathStat(pto, "confidence") <- matrix(runif(1000), ncells, npaths)
pathStatNames(pto)
```

For convenience, we also provide the `averagePseudotime()` function to compute the average pseudotime for each cell.
This is sometimes necessary in applications that only expect a single set of pseudotime values, e.g., for visualization.

```{r}
summary(averagePseudotime(pto))
```

# Session information {-}

```{r}
sessionInfo()
```
